Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-04-11 paper.Rnw
Major reorg of the methods is complete

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, url, geometry, ctable}
\usepackage{titlesec}
\usepackage[round,authoryear]{natbib}

@@

% add \subsubsubsection
\setcounter{secnumdepth}{4}  
\titleformat{\paragraph}
{\normalfont\normalsize\bfseries}{\theparagraph}{1em}{}
\titlespacing*{\paragraph}
{0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}
\newcommand{\subsubsubsection}[1]{\paragraph{#1}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         Experiment 1: Whole Hippocampus Cross-Validation            %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Experiment 1: Whole Hippocampus Cross-Validation}

To test the effect of parameter settings on the \mb algorithm, Monte
Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using a subset of
the ADNI1 dataset for validation (see below).  This form of cross-validation
allows us to validate \mb with various atlas and template library sizes,
registration and label fusion methods.

Monte Carlo Cross-Validation (MCCV), also known as repeated random sub-sampling
cross-validation, consists of repeated rounds of validation in which items from
the dataset are randomly sampled (without replacement) and assigned to a
training set or validation set \cite{Shao1993}.

In this experiment, each validation round consists of the following steps. First, 
an atlas library is selected (the training set) and, from the remaining images, 
a template library and subject to be segmented is chosen (the validation set). 
Second, \mb and basic multi-atlas segmentation are performed on the subject using the 
selected atlas and template libraries. Third, the accuracy of the resulting 
segmentations are measured against the SNT labels.

A total of ten validation rounds are perfomed on each subject in the dataset,
for each combination of parameter settings. Reported subject segmentation
accuracy is the averaged over the ten validation rounds. The parameter settings
we explore are: atlas library size (1-9), template library size (1-20),
registration method (\ants vs.  \animal), and label fusion method (MV, XC-WV,
NMI-WV). A total of $10*69*9*20*2*3 = 7452000$ validation rounds were conducted,
resulting in a total of 1'490'400 segmentations analysed.

\subsubsubsection{Dataset evaluated.}

69 1.5T images were arbitrarily selected from the {\em ADNI1:Complete 1Yr 1.5T}
standardized dataset. 23 subjects were chosen from each disease category
(cognitively normal (CN), MCI and AD). Demographics for this subset are shown in
Table \ref{tab:ADNI1-xvalidation-demographics}.

@@

The atlas and template libraries are selected from the dataset evaluated.
Atlases consist of images taken from the dataset, with corresponding manual
labels provided by SNT. Template library images were selected at random from
remaining images not used as atlases. The atlas library size ranged from 1 to 9
images. Template library size ranged from 1 to 20 images.

Both the \ants and \animal registration methods were used.

Majority vote, cross-correlation weighted majority vote, and
Normalized Mutual Information weighted majority vote were used. With the
weighted majority vote fusion methods, the number of top labels used in the
fusion was varied from 1 to 20 images.

\subsubsubsection{Evaluation method}

The Dice similarity coefficient (DSC) assesses the agreement between two
segmentations.  It is one of the most widely used measures of segmentation
performance, and we use it as the basis of comparison throughout this paper.
Additionally, we report the Jaccard index, another commonly used similarity
measure:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%        Experiment 2: Hippocampal Subfield Cross-Validation          %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Winterburn Atlases Cross-Validation}
In this experiment, the accuracy of the \mb algorithm on hippocampal subregion
segmentation is tested using a leave-one-out cross-validation (LOOCV) design.
Leave-one-out-cross-validation (LOOCV) is performed separately for each dataset
evaluated (WA-BRAVO, and WA-Resampled) as follows. Each subject in the dataset
is segmented by \mb using an atlas library in which that subject's image is
excluded.

The atlas library is composed of the Winterburn T1 atlas images and hippocampal
subregion segmentations. The template library consists of 15 3T T1 images
(0.9mm-isotropic voxels) of healthy subjects as well as all of the images from
the dataset being evaluated.

- note registration and fusion method were chosen 
\todo{reference, demographics for health control images}.

\subsubsubsection{Dataset evaluated} 
\mb segmentations are evaluated on two datasets. The WA-BRAVO dataset consists
of separately acquired 3T T1 BRAVO images (0.9mm-isotropic voxel) of four of the
five Winterburn atlas subjects.  The WA-Resampled dataset consists of the
Winterburn atlas images and segmentations downsampled to 0.9mm-isotropic voxel
resolution.

\subsubsubsection{Evaluation method}  
The segmentation volumes of each hippocampal subregion are compared to
the expert manual segmentations of the unmodified Winterburn atlases.  As a
point of comparison, we also calculate the subregion volumes from the Winterburn
atlas segmentations after downsampling to 0.9mm-isotropic voxels.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         Experiment 3: Application of \mb to the segmentation        %%%%%
%%%%%                       of schizophrenia patients                     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{SZ First Episode Patient Validation}
To validate that \mb algorithm works effectively with diseased brain images the
performance of \mb is measured on a Schizophrenia dataset.

The atlas library is composed of the Winterburn T1 atlas images, with whole
hippocampus segmentations (no subregions). The template library is composed of
21 images selected at random from the SZFEP dataset.

- note registration and fusion method were chosen 
\subsubsubsection{Dataset Evaluated}

\subsubsubsection{Evaluation method}
The manual segmentation protocol used to segment the Winterburn atlases is
similar to, but different from, the protocol used to segment the SZFEP dataset.
Therefore, rather than use an overlap metric, \mb hippocampal volumes are
compared to the corresponding manual segmentation volumes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         EXPERIMENT 3: Application of \mb to the segmentation        %%%%%
%%%%%                       of ADNI subjects                              %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 4: Segmentation of Alzheimer's Disease Patients}

To test the accuracy of \mb on a real-world task we segment the entire
ADNI-1 dataset using an atlas set that is not representative of the target set.

The Winterburn atlases were used as the atlas
library.  The template library consisted of 21 randomly selected images from the
ADNI1 data dataset (7 healthy, MCI and AD subjects).

- note registration and fusion method were chosen

\ref{tab:ADNI1-scr-tmpllib-demographics}.

\subsubsubsection{Dataset evaluated} 
All images from the {\em ADNI1:Complete 1Yr 1.5T}
standardized dataset.

\subsubsubsection{Evaluation method}
Since the hippocampal segmentation protocols differ between the ADNI labels and
Winterburn atlases, this poses a problem for direct evaluation between labels
produced by \mb and the ADNI labels in terms of overlap; we would not expect
different segmentation protocols to have a high degree of overlap.  Instead, to
evaluate the performance of \mb we compare the correlation of \mb segmentation
volumes with manual segmentation (SNT) volumes.  Additionally, we correlate the
hippocampal volumes of established automated segmentation methods to \mb
segmentations.

%%%%%                             Subjects                                 %%%%%
\subsection{Subjects}

\subsubsection{Winterburn Atlases}
The Winterburn atlases \citep{Winterburn2013} are digital hippocampal segmentations
of of five in-vivo 300u isotropic T1-weighted MR images. The segmentations
include subfield segmentations for the cornus ammonis (CA) 1, CA4, dentate
gyrus, subiculum, and CA 2 and 3 combined. Subjects in the Winterburn atlases
range in age from 29-57 years (mean age of 37), and include two males and three
females.

In addition to the high-resolution scans distributed as part of the Winterburn
atlases, we also obtained additional T1 BRAVO scans of four of the five
subjects.  \todo{demographics?}\todo{explain scan parameters in more detail..}

\subsubsubsection{FEP dataset}
All patients were recruited and treated through the Prevention and Early
Intervention Program for Psychoses (PEPP-Montreal), a specialized early
intervention service at the Douglas Mental Health University Institute in
Montreal, Canada. People aged 15–30 years from the local catchment area
suffering from either affective or non-affective psychosis who had not taken
antipsychotic medication for more than one month with an IQ above 70 were
consecutively admitted as either in- or out-patients. Of those treated at PEPP,
only patients aged 18 to 30 years with no previous history of neurological
disease or head trauma causing loss of consciousness were eligible for the
neuroimaging study; only those suffering from schizophrenia spectrum disorders
were considered for this analysis.  For complete program details see Malla et
al. \citep{Malla2003}.

Scanning was carried out at the Montreal Neurological Institute on a 1.5-T
Siemens whole body MRI system.  Structural T1 volumes were acquired for each
participant using a three-dimensional (3D) gradient echo pulse sequence with
sagittal volume excitation (repetition time=22ms, echo time=9.2ms, flip
angle=$30{\circ}$, 180 1mm contiguous sagittal slices). The rectangular
field-of-view for the images was 256mm (SI)×204mm (AP).  Subject demographics
are shown in table \ref{tab:SZFEP-Demographics}.

@@

@@

\subsubsection{ADNI1 1.5T Complete 1Yr Dataset}
Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI1 database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset download was the "ADNI1:Complete 1Yr 1.5T"
standardized dataset available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly preprocessed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008a}.  Representative 1.5T imaging parameters were TR
= 2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240
x 240mm, a 192 x 192 x 166 matrix (x, y, and z directions) yielding a voxel
resolution of 1.25 x 1.25 x 1.2 $mm^3$. Clinical and demographic data are shown
in table \ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Demographics}.

% The following blurb is taken (except for the first sentence) verbatim 
For a subset of ADNI1 images, labels of the left and right hippocampi are
available (herein refered to as SNT labels).  Semi-automated hippocampal
volumetry was carried out using a commercially available high dimensional brain
mapping tool (Medtronic Surgical Navigation Technologies, Louisville, CO), that
has previously been validated and compared to manual tracing of the hippocampus
\citep{Hsu2002}. Measurement of hippocampal volume is achieved first by placing
manually 22 control points as local landmarks for the hippocampus on the
individual brain MRI data: one landmark at the hippocampal head, one at the
tail, and four per image (i.e., at the superior, inferior, medial and lateral
boundaries) on five equally spaced images perpendicular to the long axis of the
hippocampus. Second, fluid image transformation is used to match the individual
brains to a template brain \citep{Christensen1997}. The pixels corresponding to
the hippocampus are then labeled and counted to obtain volumes. This method of
hippocampal voluming has a documented reliability of an intraclass coefficient
better than .94 \citep{Hsu2002}.

%                                                                             
%                                                                             
% ADNI1 Complete 1Yr Dataset Demographics Table 
%                                                                             
%                                                                             
<<ADNI1-scr-demographics, echo=F, results="asis",cache=FALSE>>=
# RID < 2000 ensures only ADNI1 patients
yr1 = read.csv("data/ADNI_baseline_volumes/ADNI1_Complete_1Yr_1.5T_11_15_2012.csv")
yr1$VISCODE <- factor(yr1$Visit, levels = c(1,2,3,4), labels=c("bl","m03","m06","m12"))
yr1 = subset(yr1, !grepl("Scaled_2", Description))
yr1 = yr1[,c("RID","VISCODE")]

adnimerge_yr1 = merge(adnimerge, yr1) 
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, 
               subset =  adnimerge_yr1, #RID %in% yr1$RID & VISCODE %in% c('bl','m06','m12'),   
               method = "reverse", test=FALSE, overall=TRUE)

latex(tab, file="", size="scriptsize", 
    caption="ADNI1 1.5T Complete 1Yr dataset demographics",
    label="tab:ADN1-1.5T-Complete-1Yr-Dataset-Demographics")
@ %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

>>>>

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{algorithm, color, algpseudocode, amsmath, url, geometry, ctable}
\usepackage{titlesec, siunitx}
\usepackage[round,authoryear]{natbib}

@@

% add \subsubsubsection
% \setcounter{secnumdepth}{4}  
% \titleformat{\paragraph}
% {\normalfont\normalsize\bfseries}{\theparagraph}{1em}{}
% \titlespacing*{\paragraph}
% {0pt}{3.25ex plus 1ex minus .2ex}{1.5ex plus .2ex}
\newcommand{\subsubsubsection}[1]{\paragraph{#1}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         Experiment 1: Whole Hippocampus Cross-Validation            %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 1: Whole Hippocampus Cross-Validation}

Monte Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using a pool
of images and manual hippocampal segmentations from ADNI1 dataset.  This form of
cross-validation allows us to rigorously validate a large number of parameter
settings of \mb.

\paragraph{ADNI1 dataset}

69 1.5T images were arbitrarily selected from the baseline scans in the {\em
ADNI1:Complete 1Yr 1.5T} standardized dataset. 23 subjects were chosen from each
disease category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Manual segmentations of the left and
right whole hippocampi are available. These labels have been generated using a
semiautomatic tool (SNT, Medtronic Surgical Navigation Technologies, Louisville,
CO; see Supplementary Materials).

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI1 database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset download was the "ADNI1:Complete 1Yr 1.5T"
standardized dataset available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly preprocessed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008a}.  Representative 1.5T imaging parameters were TR
= 2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240
x 240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel resolution of $1.25 \times 1.25 \times 1.2mm^3$.

@@

\paragraph{Experiment details}

Monte Carlo Cross-Validation (MCCV), also known as repeated random sub-sampling
cross-validation, consists of repeated rounds of validation conducted on a fixed
dataset \cite{Shao1993}. In each round, the dataset is randomly partitioned into
a training set and a validation set. The method to be validated is then given
the training data, and its output is compared with the validation set.

In this experiment, our dataset consists of 69 1.5T images and corresponding
manual segmentations taken from the ADNI1 project (see below for details). In each
validation round, the training set is used as an atlas library for both \mb and
basic multi-atlas. The remaining images are segmented by both methods, and the
resulting segmentations are compared to the manual segmentations for the images.

A total of ten validation rounds are perfomed on each subject in the dataset,
over each combination of parameter settings. The parameter settings
we explore are: atlas library size (1-9), template library size (1-20),
registration method (\ants or \animal), and label fusion method (majority vote,
cross-correlation weighted majority vote, and normalized mutual information
weighted majority vote).  A total of $10 \times 69 \times 9 \times 20 \times 2
\times 3 = \num{7452000}$ validation rounds were conducted, resulting in a total
of $\num{1490400}$ segmentations analysed.

\paragraph{Evaluation method}

The Dice similarity coefficient (DSC) assesses the agreement between two
segmentations. It is one of the most widely used measures of segmentation
performance, and we use it as the basis of comparison in this experiment.
Additionally, we report the Jaccard index, another commonly used similarity
measure:

@@

Reported subject segmentation accuracy is averaged over the ten validation
rounds for each parameter setting.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%        Experiment 2: Hippocampal Subfield Cross-Validation          %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 2: Winterburn Atlases Cross-Validation}
In this experiment, the accuracy of the \mb algorithm on hippocampal subregion
segmentation is tested using a leave-one-out cross-validation (LOOCV) design.
The optimal parameter settings found in Experiment 1 are used.

\paragraph{Winterburn Atlases dataset} 
The Winterburn atlases \citep{Winterburn2013} are digital hippocampal
segmentations of five in-vivo $300\mu$ isotropic T1-weighted MR images. The
segmentations include subfield segmentations for the cornus ammonis (CA) 1, CA4,
dentate gyrus, subiculum, and CA 2 and 3 combined. Subjects in the Winterburn
atlases range in age from 29-57 years (mean age of 37), and include two males
and three females.

In addition to the high-resolution scans distributed as part of the Winterburn
atlases, we also obtained additional 3T T1 BRAVO images (0.9mm-isotropic voxels)
of four of the five Winterburn atlas subjects.

\paragraph{Experiment details} 
Leave-one-out cross-validation (LOOCV) is an approach in which the method to be
validated is given all but one item in a dataset as training data, and the
output is compared with the left-out item. This is done, in turn, for each item
in the dataset.

In this experiment, the five $300\mu$-isotropic voxel Winterburn atlases are
used as the atlas library for \mb segmentation. LOOCV is conducted separately
for two different input datasets: the four Winterburn atlas subject BRAVO MR
images (referred to as the {\em Winterburn BRAVO} dataset), and the five
Winterburn atlas subject images resampled to 0.9mm-isotropic voxel resolution
(referred to as the {\em Winterburn Resampled} dataset. Each subject in the
dataset is segmented by \mb with that subject's image excluded from the atlas
library. The template library consists of 3T T1 images (0.9mm-isotropic
voxels) of healthy subjects in addition to the images from the dataset being
evaluated.

The optimal size of template library, registration method, and label fusion
method found in Experiment 1 are used.

\paragraph{Evaluation method}  
The segmentation volumes of each hippocampal subregion are compared to the
expert manual segmentations of the unmodified Winterburn atlases.  As a point of
comparison, we also calculate the subregion volumes from the Winterburn atlas
segmentations after downsampling to 0.9mm-isotropic voxels.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         Experiment 3: Application of \mb to the segmentation        %%%%%
%%%%%                       of schizophrenia patients                     %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 3: Application to the segmentation first episode
schizophrenia patients}

To validate that \mb algorithm works effectively with diseased brain images \mb
is applied to a dataset of Schizophrenia patient MR images. The resulting
segmentations are assessed for quality by comparison with expert manual
segmentations.

\paragraph{SZ-FEP dataset}
All patients were recruited and treated through the Prevention and Early
Intervention Program for Psychoses (PEPP-Montreal), a specialized early
intervention service at the Douglas Mental Health University Institute in
Montreal, Canada. People aged 15–30 years from the local catchment area
suffering from either affective or non-affective psychosis who had not taken
antipsychotic medication for more than one month with an IQ above 70 were
consecutively admitted as either in- or out-patients. Of those treated at PEPP,
only patients aged 18 to 30 years with no previous history of neurological
disease or head trauma causing loss of consciousness were eligible for the
neuroimaging study; only those suffering from schizophrenia spectrum disorders
were considered for this analysis.  For complete program details see Malla et
al. \citep{Malla2003}.

Scanning of 81 subjects was carried out at the Montreal Neurological Institute
on a 1.5-T Siemens whole body MRI system.  Structural T1 volumes were acquired
for each participant using a three-dimensional (3D) gradient echo pulse sequence
with sagittal volume excitation (repetition time=22ms, echo time=9.2ms, flip
angle=$30{\circ}$, 180 1mm contiguous sagittal slices). The rectangular
field-of-view for the images was 256mm (SI)$\times$204mm (AP).  Subject
demographics are shown in table \ref{tab:SZFEP-Demographics}.

@@

\paragraph{Experiment details} 
\mb is configured with an atlas library composed of the Winterburn T1 atlases
(see Experiment 2).  All images from the SZ-FEP dataset are segmented.  The optimal
size of template library, registration method, and label fusion method found in
Experiment 1 are used.

\paragraph{Evaluation method}
The manual segmentation protocol used to segment the Winterburn atlases is
similar to, but different from, the protocol used to segment the SZFEP dataset.
Therefore, rather than use an overlap metric, \mb hippocampal volumes are
compared to the corresponding manual segmentation volumes.

\todo{discuss Bland-Altman?  Or bias analysis}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%         EXPERIMENT 3: Application of \mb to the segmentation        %%%%%
%%%%%                       of ADNI subjects                              %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Experiment 4: Application to the segmentation of Alzheimer's
disease patients}

To validate that \mb algorithm works as well as established automated methods,
\mb is applied to the ADNI1 dataset and the resulting segmentations are
compared to those produced by FreeSurfer, FSL, MAPER, and by expert manual
segmentation.

\paragraph{ADNI1 dataset revisited} 
All images from the {\em ADNI1:Complete 1Yr 1.5T} standardized dataset described
in Experiment 1 are used.  Clinical and demographic data are shown in table
\ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Demographics}.

\paragraph{Experiment details} 
\mb is configured with an atlas library composed of the Winterburn T1 atlases
(see Experiment 2). All images from the ADN11 Complete 1Yr 1.5T dataset are
segmented. The optimal size of template library, registration method, and label
fusion method found in Experiment 1 are used. The template library is composed
of equal numbers of images from each disease class (AD, MCI, and cognitively
normal controls).

\paragraph{Evaluation method}
Since the hippocampal segmentation protocols differ between the ADNI labels and
Winterburn atlases, this poses a problem for direct evaluation between labels
produced by \mb and the ADNI labels in terms of overlap; we would not expect
different segmentation protocols to have a high degree of overlap.  Instead, to
evaluate the performance of \mb we compare the correlation of \mb segmentation
volumes with manual segmentation (SNT) volumes. Additionally, we correlate the
hippocampal volumes of established automated segmentation methods to \mb
segmentations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                                                             
% ADNI1 Complete 1Yr Dataset Demographics Table 
%                                                                             
<<ADNI1-scr-demographics, echo=F, results="asis",cache=FALSE>>=
# RID < 2000 ensures only ADNI1 patients
yr1 = read.csv("data/ADNI_baseline_volumes/ADNI1_Complete_1Yr_1.5T_11_15_2012.csv")
yr1$VISCODE <- factor(yr1$Visit, levels = c(1,2,3,4), labels=c("bl","m03","m06","m12"))
yr1 = subset(yr1, !grepl("Scaled_2", Description))
yr1 = yr1[,c("RID","VISCODE")]

adnimerge.yr1 = merge(adnimerge, yr1) 
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge.yr1, 
               #subset = RID %in% yr1$RID & VISCODE %in% c('bl','m06','m12'),   
               method = "reverse", test=FALSE, overall=TRUE)

latex(tab, file="", size="scriptsize", 
    caption="ADNI1 1.5T Complete 1Yr dataset demographics",
    label="tab:ADN1-1.5T-Complete-1Yr-Dataset-Demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\section{Supplementary Materials}

\subsection{ADNI Manual Labels}
% The following blurb is taken (except for the first sentence) verbatim 
Semi-automated hippocampal volumetry was carried out using a commercially
available high dimensional brain mapping tool (Medtronic Surgical Navigation
Technologies, Louisville, CO), that has previously been validated and compared
to manual tracing of the hippocampus \citep{Hsu2002}. Measurement of hippocampal
volume is achieved first by placing manually 22 control points as local
landmarks for the hippocampus on the individual brain MRI data: one landmark at
the hippocampal head, one at the tail, and four per image (i.e., at the
superior, inferior, medial and lateral boundaries) on five equally spaced images
perpendicular to the long axis of the hippocampus. Second, fluid image
transformation is used to match the individual brains to a template brain
\citep{Christensen1997}. The pixels corresponding to the hippocampus are then
labeled and counted to obtain volumes. This method of hippocampal voluming has a
documented reliability of an intraclass coefficient better than .94
\citep{Hsu2002}.
